### 2014-12-26 mcmc.tex
notes and content

% to-do
% -----
% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

% style notes
% -----------
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?

@@

\footnotetext{%
    The notes begin on page~\pageref{note:first}, including the
    license\note{\label{note:first}
        Copyright 2013 by the authors. This work is licensed under a
        \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
            Creative Commons Attribution-NonCommercial-NoDerivs 3.0 Unported
            License}.}
    and the acknowledgements\note{%
        We would like to thank
          Jo Bovy (IAS),
          Brendon Brewer (Auckland),
          Jonathan Goodman (NYU),
          Fengji Hou (NYU), and
          Dustin Lang (CMU)
        for valuable advice and comments.}.
}

\noindent
Dan~Foreman-Mackey\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New York University}\\[1ex]
David~W.~Hogg\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New York University}\\
\affil{Max-Planck-Institut f\"ur Astronomie, Heidelberg}

@@

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  We don't.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $x$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(x) \quad \mbox{for all $x$}
\\
1 &=& \int p(x)\,\dd x
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $x$.
Importantly, although $p(x)$ is single-valued, $x$ can be a scalar or
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $x$ will often be the full blob of free
parameters in the model.

@@

\section{What is M--H MCMC?}

@@

>>>>

% This file is part of the MCMC project,
%   which is part of the Data Analysis Recipes project
% Copyright 2013, 2014 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).

% to-do
% -----
% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
%   - Hill
% - post on arXiv

% style notes
% -----------
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
% - \note{} right after word if you are endnoting a phrase,
%   but after the full stop if endnoting the sentence or paragraph.

@@

\footnotetext{%
    The notes begin on page~\pageref{note:first}, including the
    license\note{\label{note:first}
        Copyright 2013, 2014 by the authors. This work is licensed under a
        \href{http://creativecommons.org/licenses/by-nc-nd/3.0/deed.en\_US}{%
            Creative Commons Attribution-NonCommercial-NoDerivs 3.0 Unported
            License}.}
    and the acknowledgements\note{%
        We would like to thank
          Jo Bovy (IAS),
          Brendon Brewer (Auckland),
          Jonathan Goodman (NYU),
          Fengji Hou (NYU), and
          Dustin Lang (CMU)
        for valuable advice and comments.}.
}

\noindent
Dan~Foreman-Mackey\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\[1ex]
David~W.~Hogg\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\
\affil{Center~for~Data~Science,
       New~York~University}\\
\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}

@@

Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  We don't.  It has something to do with measure theory.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $x$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(x) \quad \mbox{for all $x$}
\\ \label{eq:normp}
1 &=& \int p(x)\,\dd x
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $x$.
Importantly, although $p(x)$ is single-valued, $x$ can be a scalar or
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $x$ will often be the full blob of free
parameters in the model.
Implicitly, the integral in \equationname~(\ref{eq:normp}) is high-dimensional; it
has as many dimensions as there are elements or entries or components of $x$.
Also, if there are elements or entries or components of $x$ that are discrete
(that is, take on only integer values or equivalent),
then along those dimensions the integral becomes a discrete sum.
This latter is a detail to which we return below.

@@

\section{What is Metropolis--Hastings MCMC?}

The simplest algorithm for MCMC is the Metropolis--Hastings algorithm.\note{%
  HOGG put here some short discussion of the name of the algorithm and the original reference.}
It is so simple, that we recommend that any reader of this document
who has not previously implemented the algorithm take a break at the end
of this \sectionname\ and implement it forthwith, in a short piece of computer code,
in the context of some simple problem.\note{%
  HOGG put reference to fitting a line here and point inside to the simple problem suggestion.}

@@

HOGG:  How to deal with discrete or integer or binary parameters?
